HTW Modeling

### 2024-03-27 Model/htw_model.qmd
add indv sbj plot, and final e2 e3 results paragraph

---
title: HTW Modeling 
author: Thomas Gorman
#date:  #"`r Sys.Date()`"
page-layout: full
categories: [Modeling, ALM, EXAM, R]
lightbox: true
toc: false
format-links: false
code-fold: true
code-tools: true
execute: 
  warning: false
  eval: true
# format:
#   html: default
#   hugo-md:
#     echo: false
#     html-math-method: mathjax
#     output-file: e1-hugo.md
  # gfm:
  #   echo: true
  #   output-file: e1-gfm.md
prefer-html: true
---

```{r}
pacman::p_load(dplyr,purrr,tidyr,ggplot2, data.table, here, patchwork, conflicted, 
               stringr,future,furrr, knitr, reactable, flextable,ggstance, htmltools,ggdist,ggh4x,brms,tidybayes,bayestestR, gt)
#conflict_prefer_all("dplyr", quiet = TRUE)
walk(c("flextable","dplyr"), conflict_prefer_all, quiet = TRUE)
#options(brms.backend="cmdstanr",mc.cores=4)
options(digits=3, scipen=999, dplyr.summarise.inform=FALSE)
walk(c("Display_Functions","fun_alm","fun_indv_fit","fun_model", "prep_model_data","org_functions"), ~source(here::here(paste0("Functions/", .x, ".R"))))

@@

```{r}
#| cache: false

@@

In project 1, I applied model-based techniques to quantify and control for the similarity between training and testing experience, which in turn enabled us to account for the difference between varied and constant training via an extended version of a similarity based generalization model. In project 2, I will go a step further, implementing a full process model capable of both 1) producing novel responses and 2) modeling behavior in both the learning and testing stages of the experiment. For this purpose, we will apply the associative learning model (ALM) and the EXAM model of function learning [@deloshExtrapolationSineQua1997]. ALM is a simple connectionist learning model which closely resembles Kruschke's ALCOVE model [@kruschkeALCOVEExemplarbasedConnectionist1992], with modifications to allow for the generation of continuous responses.

@@

::: {.callout}
**{{< fa regular lightbulb >}} Approximate Bayesian Computation**

@@

bct_e1 <- as.data.frame(bayestestR::describe_posterior(e1_ee_brm_ae, centrality = "Mean")) %>%
  select(1,2,4,5,6) %>%
  setNames(c("Term", "Estimate","95% CrI Lower", "95% CrI Upper", "pd")) %>%
  mutate(across(where(is.numeric), \(x) round(x, 2))) %>%
  tibble::remove_rownames() %>%
  mutate(Term = stringr::str_remove(Term, "b_")) %>%
  kable(booktabs = TRUE)

@@

bm1_e2 <- get_coef_details(e2_ee_brm_ae, "conditVaried")
bm2_e2 <- get_coef_details(e2_ee_brm_ae, "ModelEXAM")
bm3_e2 <- get_coef_details(e2_ee_brm_ae, "ModelEXAM:conditVaried")
# bct_e2 <- as.data.frame(bayestestR::describe_posterior(e2_ee_brm_ae, centrality = "Mean")) %>%
#   select(1,2,4,5,6) %>%
#   setNames(c("Term", "Estimate","95% CrI Lower", "95% CrI Upper", "pd")) %>%
#   mutate(across(where(is.numeric), \(x) round(x, 2))) %>%
#   tibble::remove_rownames() %>%
#   mutate(Term = stringr::str_remove(Term, "b_")) %>%
#   kable(booktabs = TRUE)

@@

# bct_e3  <- as.data.frame(bayestestR::describe_posterior(e3_ee_brm_ae, centrality = "Mean")) %>%
#   select(1,2,4,5,6) %>%
#   setNames(c("Term", "Estimate","95% CrI Lower", "95% CrI Upper", "pd")) %>%
#   mutate(across(where(is.numeric), \(x) round(x, 2))) %>%
#   tibble::remove_rownames() %>%
#   mutate(Term = stringr::str_remove(Term, "b_")) %>%
#   kable(booktabs = TRUE)

msa <- modelsummary::modelsummary(list("Exp 1."=e1_ee_brm_ae,"Exp 2."=e2_ee_brm_ae,"Exp 3. "=e3_ee_brm_ae), 
    gof_map = NA,
    fmt=1,
    estimate = "{estimate}",
    statistic="[{conf.low}, {conf.high}]", 
    coef_omit = "^sigma|^sd|^cor",
    coef_rename=function(x) stringr::str_remove(x, "b_"), output="gt")

msa

#library(modelsummary)
# modelsummary::modelsummary(get_estimates(e1_ee_brm_ae, test = c("pd")),gof_map = NA)
#modelsummary::get_estimates(e1_ee_brm_ae,test="pd")
```

@@

```{r}
#| fig-height: 8
#| fig-width: 10
#| fig-cap: "Empirical data and Model predictions from Experiment 2 and 3 for the testing stage. Observed data is shown on the right. Bolded bars indicate bands that were trained, non-bold bars indicate extrapolation bands.  "

rbind(e2_model$post_dat_l |> filter( Fit_Method=="Test_Train") |> 
  group_by(id,condit, Fit_Method,Resp,bandType,x,vb) |> 
 summarize(vx=median(val)) |> mutate(Exp="E2",bandOrder="Reverse"), 
 e3_model$post_dat_l |> filter( Fit_Method=="Test_Train") |> 
  group_by(id,condit, Fit_Method,Resp,bandType,x,vb,bandOrder) |>
  summarize(vx=median(val)) |> mutate(Exp="E3")) |>
  ggplot( aes(x=condit,y=vx, fill=vb,col=ifelse(bandType=="Trained","black",NA),size=ifelse(bandType=="Trained","black",NA))) +
  stat_bar + 
    facet_nested_wrap(~Exp+bandOrder+Resp, strip.position = "top", scales = "free_x") +
    scale_color_manual(values = c("black" = "black"), guide = "none") +
  scale_size_manual(values = c("black" = .7), guide = "none") +
    theme(panel.spacing = unit(0, "lines"), 
        #  strip.background = element_blank(),
        #  strip.placement = "outside",
         legend.position = "none",plot.title = element_text(hjust=.50),
         axis.title.x = element_blank(),
         plot.margin = unit(c(20,0,0,0), "pt")) + 
         labs(title="Model Predictions Experiment 2 & 3", y="vx")

```{r}
#| label: fig-htw-best-model
#| fig-cap: Difference in model errors for each participant, with models fit to both train and test data. Positive values favor EXAM, while negative values favor ALM. 
#| fig-height: 12
#| fig-width: 11

@@

```{r  fig.width=11, fig.height=10}
#| label: fig-htw-indv-pred
#| fig-cap: Model predictions alongside observed data for a subset of individual participants. A) 3 constant and 3 varied participants fit to both the test and training data. B) 3 constant and 3 varied subjects fit to only the trainign data. Bolded bars indicate bands that were trained, non-bold bars indicate extrapolation bands. 
#| fig-height: 11
#| fig-width: 12

cId_tr <- c(137, 181, 11)
vId_tr <- c(14, 193, 47)
cId_tt <- c(11, 93, 35)
vId_tt <- c(1,14,74)
cId_new <- c(175, 185, 134, 155, 68, 93, 74, 134)
# filter(id %in% (filter(bestTestEXAM,group_rank<=9, Fit_Method=="Test")

# e3 sbjs - 245, 219, 264, 280, 249, 274

testIndv <- post_dat_l |> filter(id %in% c(cId_tt,vId_tt,cId_new), Fit_Method=="Test_Train") |> 
   mutate(x=as.factor(x), Resp=as.factor(Resp)) |>
  group_by(id,condit,Fit_Method,Model,Resp) |>
   mutate(flab=paste0("Subject: ",id)) |>
  ggplot(aes(x = Resp, y = val, fill=vb, col=ifelse(bandType=="Trained","black",NA),size=ifelse(bandType=="Trained","black",NA))) + 
  stat_bar_sd + ggh4x::facet_nested_wrap(condit~flab, axes = "all",ncol=3) +
  scale_color_manual(values = c("black" = "black"), guide = FALSE) +
  scale_size_manual(values = c("black" = .5), guide = FALSE) + 
  labs(title="Individual Participant fits from Test & Train Fitting Method",
       y="X Velocity",fill="Target Velocity") +
   guides(fill = guide_legend(nrow = 1)) + 
  theme(legend.position = "bottom",axis.title.x = element_blank())

>>>>

---
title: HTW Modeling 
author: Thomas Gorman
#date:  #"`r Sys.Date()`"
page-layout: full
categories: [Modeling, ALM, EXAM, R]
lightbox: true
toc: false
cache: false
format-links: false
code-fold: true
code-tools: true
execute: 
  warning: false
  eval: true
format:
  html: default
  hugo-md:
    echo: false
    html-math-method: mathjax
    output-file: e1-hugo.md
  # gfm:
  #   echo: true
  #   output-file: e1-gfm.md
prefer-html: true
---

```{r}
#| cache: false
pacman::p_load(dplyr,purrr,tidyr,ggplot2, data.table, here, patchwork, conflicted, 
               stringr,future,furrr, knitr, reactable, flextable,ggstance, htmltools,ggdist,ggh4x,brms,tidybayes,bayestestR, gt)
#conflict_prefer_all("dplyr", quiet = TRUE)
walk(c("flextable","dplyr"), conflict_prefer_all, quiet = TRUE)
#options(brms.backend="cmdstanr",mc.cores=4)
options(digits=3, scipen=999, dplyr.summarise.inform=FALSE)
walk(c("Display_Functions","fun_alm","fun_indv_fit","fun_model", "prep_model_data","org_functions"), ~source(here::here(paste0("Functions/", .x, ".R"))))

@@

```{r}
#| cache: true

@@

The modeling goal is to implement a full process model capable of both 1) producing novel responses and 2) modeling behavior in both the learning and testing stages of the experiment. For this purpose, we will apply the associative learning model (ALM) and the EXAM model of function learning [@deloshExtrapolationSineQua1997]. ALM is a simple connectionist learning model which closely resembles Kruschke's ALCOVE model [@kruschkeALCOVEExemplarbasedConnectionist1992], with modifications to allow for the generation of continuous responses.

@@

::: {.callout}

**{{< fa regular lightbulb >}} Approximate Bayesian Computation**

@@

bct_e1 <- as.data.frame(bayestestR::describe_posterior(e1_ee_brm_ae, centrality = "Mean")) %>%
  select(1,2,4,5,6) %>%
  setNames(c("Term", "Estimate","95% CrI Lower", "95% CrI Upper", "pd")) %>%
  mutate(across(where(is.numeric), \(x) round(x, 2))) %>%
  tibble::remove_rownames() %>%
  mutate(Term = stringr::str_remove(Term, "b_")) #%>% kable(booktabs = TRUE)

@@

bm1_e2 <- get_coef_details(e2_ee_brm_ae, "conditVaried")
bm2_e2 <- get_coef_details(e2_ee_brm_ae, "ModelEXAM")
bm3_e2 <- get_coef_details(e2_ee_brm_ae, "ModelEXAM:conditVaried")
bct_e2 <- as.data.frame(bayestestR::describe_posterior(e2_ee_brm_ae, centrality = "Mean")) %>%
  select(1,2,4,5,6) %>%
  setNames(c("Term", "Estimate","95% CrI Lower", "95% CrI Upper", "pd")) %>%
  mutate(across(where(is.numeric), \(x) round(x, 2))) %>%
  tibble::remove_rownames() %>%
  mutate(Term = stringr::str_remove(Term, "b_")) # %>% kable(booktabs = TRUE)

@@

bct_e3  <- as.data.frame(bayestestR::describe_posterior(e3_ee_brm_ae, centrality = "Mean")) %>%
  select(1,2,4,5,6) %>%
  setNames(c("Term", "Estimate","95% CrI Lower", "95% CrI Upper", "pd")) %>%
  mutate(across(where(is.numeric), \(x) round(x, 2))) %>%
  tibble::remove_rownames() %>%
  mutate(Term = stringr::str_remove(Term, "b_")) #%>% kable(booktabs = TRUE)

# msa <- modelsummary::modelsummary(list("Exp 1."=e1_ee_brm_ae,"Exp 2."=e2_ee_brm_ae,"Exp 3. "=e3_ee_brm_ae), 
#     gof_map = NA,
#     fmt=1,
#     estimate = "{estimate}",
#     statistic="[{conf.low}, {conf.high}]", 
#     coef_omit = "^sigma|^sd|^cor",
#     coef_rename=function(x) stringr::str_remove(x, "b_"), output="gt")
#msa

bct <- rbind(bct_e1 |> mutate(exp="Exp 1"),bct_e2 |> mutate(exp= "Exp 2"),bct_e3 |> mutate(exp="Exp 3")) |> relocate(exp, .before=Term)

bct_table <- bct %>%
  mutate(
    across(c(Estimate, `95% CrI Lower`, `95% CrI Upper`), ~ round(., 2)),
    pd = round(pd, 2)
  ) %>%
  gt() %>%
  # tab_header(
  #   title = "Bayesian Model Results",
  #   subtitle = "Estimates and Credible Intervals for Each Term Across Experiments"
  # ) %>%
  cols_label(
    exp = "Experiment",
    Term = "Term",
    Estimate = "Estimate",
    `95% CrI Lower` = "95% CrI Lower",
    `95% CrI Upper` = "95% CrI Upper",
    pd = "pd"
  ) %>%
  fmt_number(
    columns = c(Estimate, `95% CrI Lower`, `95% CrI Upper`),
    decimals = 2
  ) %>%
  fmt_number(
    columns = pd,
    decimals = 2
  ) %>%
  tab_spanner(
    label = "Credible Interval",
    columns = c(`95% CrI Lower`, `95% CrI Upper`)
  ) %>%
  tab_style(
    style = list(
      #cell_fill(color = "lightgray"),
      cell_text(weight = "bold"), 
      cell_fill(color = "white"),
      cell_borders(sides = "top", color = "black", weight = px(1))
    ),
    locations = cells_body(
      columns = c(Estimate, pd),
      rows = Term=="ModelEXAM:conditVaried"
    )
  ) %>%
  # tab_style(
  #   style = list(
  #     cell_fill(color = "lightblue"),
  #     cell_text(style = "italic")
  #   ),
  #   locations = cells_body(
  #     columns = exp,
  #     rows = Term == "Intercept"
  #   )
  # ) %>%
   tab_row_group(
    label = "Experiment 3",
    rows = exp == "Exp 3"
  ) %>%
  tab_row_group(
    label = "Experiment 2",
    rows = exp == "Exp 2"
  ) %>%
  tab_row_group(
    label = "Experiment 1",
    rows = exp == "Exp 1"
  ) %>%
  tab_options(
    table.font.size = 10,
    heading.title.font.size = 16,
    heading.subtitle.font.size = 14,
    quarto.disable_processing = TRUE
    #row_group.background.color = "gray95"
  )
bct_table

*Model Fits to Experiment 2 and 3.*  The model fitting results for Experiments 2 and 3 closely mirrored those observed in Experiment 1. The Bayesian regression models predicting model error as a function of Model (ALM vs. EXAM), Condition (Constant vs. Varied), and their interaction (see @tbl-htw-ee-e23) revealed a consistent main effect of Model across all three experiments. The negative coefficients for the ModelEXAM term (Exp 2: $\beta$ = `r bm1_e2$estimate`, 95% CrI `r bm1_e2$conf.low`, `r bm1_e2$conf.high`, pd = `r bm1_e2$pd`; Exp 3: $\beta$ = `r bm1_e3$estimate`, 95% CrI `r bm1_e3$conf.low`, `r bm1_e3$conf.high`, pd = `r bm1_e3$pd`) indicate that EXAM outperformed ALM in both experiments. Furthermore, the interaction between Model and Condition was significant in both Experiment 2 ($\beta$ = `r bm3_e2$estimate`, 95% CrI `r bm3_e2$conf.low`, `r bm3_e2$conf.high`, pd = `r bm3_e2$pd`) and Experiment 3 ($\beta$ = `r bm3_e3$estimate`, 95% CrI `r bm3_e3$conf.low`, `r bm3_e3$conf.high`, pd = `r bm3_e3$pd`), suggesting that the superiority of EXAM over ALM was more pronounced for the Constant group compared to the Varied group, as was the case in Experiment 1. One notable difference in Experiment 3 was the inclusion of the bandOrder term and its interactions, which accounted for the effect of the order in which the velocity bands were presented during training and testing. Despite this additional factor, the overall pattern of results remained consistent with the previous experiments, providing strong support for the robustness of EXAM's performance, particularly in the Constant condition, across all three studies.

```{r}

@@

```{r}
#| label: fig-cm-vx-pat-e2-e3
#| fig-height: 8
#| fig-width: 10
#| fig-cap: "Empirical data and Model predictions from Experiment 2 and 3 for the testing stage. Observed data is shown on the right. Bolded bars indicate bands that were trained, non-bold bars indicate extrapolation bands."

rbind(e2_model$post_dat_l |> filter( Fit_Method=="Test_Train") |> 
  group_by(id,condit, Fit_Method,Resp,bandType,x,vb) |> 
 summarize(vx=median(val)) |> mutate(Exp="E2",bandOrder="Reverse"), 
 e3_model$post_dat_l |> filter( Fit_Method=="Test_Train") |> 
  group_by(id,condit, Fit_Method,Resp,bandType,x,vb,bandOrder) |>
  summarize(vx=median(val)) |> mutate(Exp="E3")) |>
  ggplot( aes(x=condit,y=vx, fill=vb,col=ifelse(bandType=="Trained","black",NA),size=ifelse(bandType=="Trained","black",NA))) +
  stat_bar + 
    facet_nested_wrap(~Exp+bandOrder+Resp, strip.position = "top", scales = "free_x") +
    scale_color_manual(values = c("black" = "black"), guide = "none") +
  scale_size_manual(values = c("black" = .7), guide = "none") +
    theme(panel.spacing = unit(0, "lines"), 
        #  strip.background = element_blank(),
        #  strip.placement = "outside",
         legend.position = "none",plot.title = element_text(hjust=.50),
         axis.title.x = element_blank(),
         plot.margin = unit(c(20,0,0,0), "pt")) + 
         labs(title="Model Predictions Experiment 2 & 3", y="vx")
```

```{r}
#| label: fig-htw-best-model
#| fig-cap: "Difference in model errors for each participant, with models fit to both train and test data. Positive values favor EXAM, while negative values favor ALM." 
#| fig-height: 12
#| fig-width: 11

@@

```{r  fig.width=11, fig.height=10}
#| label: fig-htw-indv-pred
#| fig-cap: Model predictions alongside observed data for a subset of individual participants. A) 3 constant and 3 varied participants fit to both the test and training data. B) 3 constant and 3 varied subjects fit to only the trainign data. Bolded bars indicate bands that were trained, non-bold bars indicate extrapolation bands. 
#| fig-height: 12
#| fig-width: 12

cId_tr <- c(137, 181, 11)
vId_tr <- c(14, 193, 47)
cId_tt <- c(11, 93, 35)
vId_tt <- c(1,14,74)
cId_new <- c(175, 134, 155, 68, 93, 74, 134)
# filter(id %in% (filter(bestTestEXAM,group_rank<=9, Fit_Method=="Test")

 e3_sbjs <-  c(245, 219, 264, 280, 249, 274)
e2_sbjs <- c(197, 157, 312, 334)

indv_post_l <- post_dat_l  |> mutate(Exp="E1",bandOrder="Original") |> select(-signed_dist) |>
  rbind(e2_model$post_dat_l |> mutate(Exp="E2",bandOrder="Reverse")) |>
  rbind(e3_model$post_dat_l |> mutate(Exp="E3") |> select(-fb)) |>
  filter(Fit_Method=="Test_Train", id %in% c(cId_tt,vId_tt,cId_new, e3_sbjs, e2_sbjs))

testIndv <- indv_post_l |> 
#filter(id %in% c(cId_tt,vId_tt,cId_new), Fit_Method=="Test_Train") |> 
   mutate(x=as.factor(x), Resp=as.factor(Resp)) |>
  group_by(id,condit,Fit_Method,Model,Resp) |>
   mutate(flab=paste0("Subject: ",id)) |>
  ggplot(aes(x = Resp, y = val, fill=vb, col=ifelse(bandType=="Trained","black",NA),size=ifelse(bandType=="Trained","black",NA))) + 
  stat_bar_sd + 
  ggh4x::facet_nested_wrap(condit~flab, axes = "all",ncol=3) +
  scale_color_manual(values = c("black" = "black"), guide = FALSE) +
  scale_size_manual(values = c("black" = .5), guide = FALSE) + 
  labs(title="Individual Participant fits from Test & Train Fitting Method",
       y="X Velocity",fill="Target Velocity") +
   guides(fill = guide_legend(nrow = 1)) + 
  theme(legend.position = "bottom",axis.title.x = element_blank())
